---
title: "Superconductivity_Classification_Project"
author: "Tomas Tisovsky, Jakub Ivacek"
output: html_document
date: "2026-04-30"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE
)
library(e1071)
library(randomForest)
library(rpart)
library(rpart.plot)
library(MLmetrics)
library(tidyverse)
library(patchwork)
library(pheatmap)
library(glmnet)
library(MASS)
library(caret)
library(pROC)
library(magrittr)
select <- dplyr::select
```

# 1. Introduction

This project focuses on the classification of superconducting materials. The goal is to predict whether a material belongs to the class of high-temperature superconductors based on its physical and chemical properties.

The original target variable is `critical_temp`. For this project, the problem is transformed into a binary classification task. Materials with critical temperature greater than or equal to 77 K are assigned to the class `high_tc`, while the remaining materials are assigned to the class `non_high_tc`.

# 2. Research Question and Hypothesis

The main research question is whether high-temperature superconductors can be identified using selected numerical material properties.

The main hypothesis is that not all available features are equally informative. Therefore, feature selection methods can help identify a smaller and more interpretable subset of variables while preserving good classification performance.

# 3. Loading Data

```{r eda-load-data}
# Load the original data file.
data_raw <- readr::read_csv("train.csv", show_col_types = FALSE)
```

## Exploratory Data Analysis

```{=html}
<!--
Write this part in your own words before submission.
Suggested content:
- The goal of this EDA is to check whether the dataset is suitable for the selected classification task.
- The analysis focuses on data quality, target distribution, class balance, predictor screening, redundancy among predictors, and outliers.
- The plots are selected only when they support modelling decisions, not as an exhaustive visual dump.
-->
```

## Scenario 1: Classification methods

In the first scenario, we compare several classification methods applied to the superconductor dataset. The goal is to evaluate how different classifiers perform at identifying high-temperature superconductors, and whether the choice of feature space affects classification performance.

The compared classification methods are:

-   logistic regression (interpretable linear classifier),
-   support vector machine with radial kernel (non-linear, margin-based classifier),
-   random forest (ensemble tree-based classifier),
-   decision tree (single tree classifier, interpretable).

Two feature spaces are compared for logistic regression and SVM:

-   20 features selected by standardized mean difference during EDA,
-   9 deduplicated features - highly correlated groups are reduced to one per group.

Random forest and decision tree use the full set of available numerical features.

```{r task1-setup, message=FALSE, warning=FALSE}

set.seed(42)
load("prepared_data.RData")

task1_train   <- train_df
task1_test    <- test_df
task1_y_train <- y_train
task1_y_test  <- y_test
```

```{r task1-helpers, message=FALSE, warning=FALSE}
evaluate_clf <- function(truth, prob, threshold = 0.5) {
  pred  <- if_else(prob >= threshold, "high_tc", "non_high_tc") %>%
    factor(levels = c("non_high_tc", "high_tc"))
  truth <- factor(truth, levels = c("non_high_tc", "high_tc"))
  cm    <- confusionMatrix(pred, truth, positive = "high_tc")
  roc_obj <- pROC::roc(response = truth, predictor = prob,
                       levels = c("non_high_tc", "high_tc"), quiet = TRUE)
  tibble(
    threshold    = threshold,
    accuracy     = unname(cm$overall["Accuracy"]),
    sensitivity  = unname(cm$byClass["Sensitivity"]),
    specificity  = unname(cm$byClass["Specificity"]),
    precision    = unname(cm$byClass["Precision"]),
    f1           = unname(cm$byClass["F1"]),
    balanced_acc = unname(cm$byClass["Balanced Accuracy"]),
    auc          = as.numeric(pROC::auc(roc_obj)),
    tp           = cm$table["high_tc",     "high_tc"],
    fn           = cm$table["non_high_tc", "high_tc"],
    fp           = cm$table["high_tc",     "non_high_tc"]
  )
}

get_youden_cutoff <- function(truth, prob) {
  truth <- factor(truth, levels = c("non_high_tc", "high_tc"))
  roc_obj <- pROC::roc(response = truth, predictor = prob,
                       levels = c("non_high_tc", "high_tc"), quiet = TRUE)
  pROC::coords(roc_obj, x = "best", best.method = "youden",
               ret = "threshold", transpose = FALSE) %>%
    as.numeric() %>% first()
}
```

### Logistic Regression

Logistic regression is the most interpretable model in this scenario. Each fitted coefficient directly represents the log-odds change increase per unit in the corresponding feature. p-values and confidence intervals provide formal statistical tests for each predictor. Two feature spaces are evaluated to assess how collinearity affects logistic regression performance and interpretability.

#### LR-A: 20 EDA-selected features

The first logistic regression model uses the top 20 features ranked by standardized mean difference (SMD). SMD measures how well a predictor separates the two classes on average, giving a model-free ranking. This feature space provides maximum signal but includes several highly correlated predictors.

```{r task1-lr-a-fit, message=FALSE, warning=FALSE}
model_lr_a <- glm(
  tc_class ~ .,
  data   = bind_cols(task1_train %>% select(all_of(top20_eda)), tc_class = task1_y_train),
  family = binomial()
)

cat("LR-A converged:", model_lr_a$converged, "\n")
cat("LR-A features:", length(top20_eda), "\n")
```

```{r task1-lr-a-eval, message=FALSE, warning=FALSE}
prob_lr_a <- predict(model_lr_a,
                     newdata = task1_test %>% select(all_of(top20_eda)),
                     type = "response")

lr_a_metrics_05 <- evaluate_clf(task1_y_test, prob_lr_a, threshold = 0.5) %>%
  mutate(threshold_type = "0.5")

youden_lr_a <- get_youden_cutoff(task1_y_test, prob_lr_a)

lr_a_metrics_youden <- evaluate_clf(task1_y_test, prob_lr_a, threshold = youden_lr_a) %>%
  mutate(threshold_type = "Youden")

lr_a_metrics <- bind_rows(lr_a_metrics_05, lr_a_metrics_youden) %>%
  mutate(model = "LR-A (20 feat)") %>%
  dplyr::select(model, threshold_type, everything())

lr_a_metrics
```

```{r task1-lr-a-roc, message=FALSE, warning=FALSE, fig.width=6, fig.height=5}
roc_lr_a <- pROC::roc(task1_y_test, prob_lr_a,
                      levels = c("non_high_tc", "high_tc"), quiet = TRUE)
plot(roc_lr_a,
     main = paste0("ROC — LR-A (20 features)  AUC = ", round(auc(roc_lr_a), 3)))
youden_coords_a <- pROC::coords(roc_lr_a, x = "best", best.method = "youden",
                                ret = c("specificity", "sensitivity"), transpose = FALSE)
points(youden_coords_a$specificity, youden_coords_a$sensitivity,
       pch = 19, col = "red", cex = 1.5)
legend("bottomright", legend = "Youden threshold", col = "red", pch = 19, bty = "n")
```

With the default threshold of 0.5, LR-A achieved an accuracy of 0.871 and sensitivity of 0.642, meaning that about 36% of true high-temperature superconductors were missed. The Youden threshold lowered the decision boundary and increased sensitivity to 0.946, recovering the majority of missed superconductors at the cost of lower specificity. However, two features (mean_Valence, gmean_Valence) showed non-significant coefficients — a sign of collinearity inflating standard errors.

#### LR-B: 9 deduplicated features

The second logistic regression model uses a reduced set of 9 features obtained by removing redundant predictors from the top-20 set. Collinearity groups were identified from the correlation matrix, and one representative per group was retained.

```{r task1-lr-b-fit, message=FALSE, warning=FALSE}
cat("LR-B features:", length(lr2_features), "\n")
cat("Features:", paste(lr2_features, collapse = ", "), "\n")

model_lr_b <- glm(
  tc_class ~ .,
  data   = bind_cols(task1_train %>% select(all_of(lr2_features)), tc_class = task1_y_train),
  family = binomial()
)

cat("LR-B converged:", model_lr_b$converged, "\n")
```

```{r task1-lr-b-eval, message=FALSE, warning=FALSE}
prob_lr_b <- predict(model_lr_b,
                     newdata = task1_test %>% select(all_of(lr2_features)),
                     type = "response")

lr_b_metrics_05 <- evaluate_clf(task1_y_test, prob_lr_b, threshold = 0.5) %>%
  mutate(threshold_type = "0.5")

youden_lr_b <- get_youden_cutoff(task1_y_test, prob_lr_b)

lr_b_metrics_youden <- evaluate_clf(task1_y_test, prob_lr_b, threshold = youden_lr_b) %>%
  mutate(threshold_type = "Youden")

lr_b_metrics <- bind_rows(lr_b_metrics_05, lr_b_metrics_youden) %>%
  mutate(model = "LR-B (9 feat)") %>%
  dplyr::select(model, threshold_type, everything())

lr_b_metrics
```

```{r task1-lr-b-roc, message=FALSE, warning=FALSE, fig.width=6, fig.height=5}
roc_lr_b <- pROC::roc(task1_y_test, prob_lr_b,
                      levels = c("non_high_tc", "high_tc"), quiet = TRUE)
plot(roc_lr_b,
     main = paste0("ROC — LR-B (9 features)  AUC = ", round(auc(roc_lr_b), 3)))
youden_coords_b <- pROC::coords(roc_lr_b, x = "best", best.method = "youden",
                                ret = c("specificity", "sensitivity"), transpose = FALSE)
points(youden_coords_b$specificity, youden_coords_b$sensitivity,
       pch = 19, col = "red", cex = 1.5)
legend("bottomright", legend = "Youden threshold", col = "red", pch = 19, bty = "n")
```

With 9 deduplicated features, LR-B is more stable no inflated standard errors than LR-A — all coefficients are statistically significant. The Youden threshold achieved the highest sensitivity among all tested models (0.970), meaning only 24 true superconductors were missed. This came at the cost of a low specificity of 0.764, producing a large number of false positives.

#### Comparison of LR variants

```{r task1-lr-comparison, message=FALSE, warning=FALSE}
lr_comparison <- bind_rows(lr_a_metrics, lr_b_metrics) %>%
  dplyr::select(model, threshold_type, threshold, accuracy, sensitivity,
                specificity, precision, f1, balanced_acc, auc)

lr_comparison
```

LR-A achieves slightly higher AUC (0.928 vs 0.921) because the 20-feature set contains more signal despite the redundancy. However, the non-significant predictors in LR-A reduce its interpretability. LR-B with the Youden threshold has the highest sensitivity (0.970) and the fewest missed superconductors (24), making it the winner in terms of senstivity. However, its specificity is lower and gives us 774 false positives which could be a huge time and resource problem in laboratory testing.

### Support Vector Machine

Support vector machine with a radial basis function (RBF) kernel finds a non-linear decision boundary by maximising the margin between classes in a high-dimensional feature space. Because SVM is sensitive to feature scale, all predictors are centred and scaled using parameters estimated from the training set only.

Two feature spaces are evaluated like in LR to asses how collinearity of features affects SVM performance.

#### SVM-A: 20 EDA-selected features

```{r task1-svm-a-fit, message=FALSE, warning=FALSE}
preproc_a          <- preProcess(task1_train %>% select(all_of(top20_eda)),
                                 method = c("center", "scale"))
train_svm_a_scaled <- predict(preproc_a, task1_train %>% select(all_of(top20_eda)))
test_svm_a_scaled  <- predict(preproc_a, task1_test  %>% select(all_of(top20_eda)))

model_svm_a <- svm(
  x           = train_svm_a_scaled,
  y           = task1_y_train,
  kernel      = "radial",
  probability = TRUE
)
```

```{r task1-svm-a-eval, message=FALSE, warning=FALSE}
prob_svm_a <- attr(
  predict(model_svm_a, test_svm_a_scaled, probability = TRUE),
  "probabilities"
)[, "high_tc"]

svm_a_metrics_05 <- evaluate_clf(task1_y_test, prob_svm_a, threshold = 0.5) %>%
  mutate(threshold_type = "0.5")

youden_svm_a <- get_youden_cutoff(task1_y_test, prob_svm_a)

svm_a_metrics_youden <- evaluate_clf(task1_y_test, prob_svm_a, threshold = youden_svm_a) %>%
  mutate(threshold_type = "Youden")

svm_a_metrics <- bind_rows(svm_a_metrics_05, svm_a_metrics_youden) %>%
  mutate(model = "SVM-A (20 feat)") %>%
  dplyr::select(model, threshold_type, everything())

svm_a_metrics
```

```{r task1-svm-a-roc, message=FALSE, warning=FALSE, fig.width=6, fig.height=5}
roc_svm_a <- pROC::roc(task1_y_test, prob_svm_a,
                       levels = c("non_high_tc", "high_tc"), quiet = TRUE)
plot(roc_svm_a,
     main = paste0("ROC — SVM-A (20 features)  AUC = ", round(auc(roc_svm_a), 3)))
youden_coords_svm_a <- pROC::coords(roc_svm_a, x = "best", best.method = "youden",
                                    ret = c("specificity", "sensitivity"), transpose = FALSE)
points(youden_coords_svm_a$specificity, youden_coords_svm_a$sensitivity,
       pch = 19, col = "red", cex = 1.5)
legend("bottomright", legend = "Youden threshold", col = "red", pch = 19, bty = "n")
```

With 20 features, SVM-A contains several correlated predictors. These duplicate dimensions in the kernel space and can distort the margin, which limits performance compared to the reduced feature set as we will see in SVM-B.

#### SVM-B: 9 deduplicated features

```{r task1-svm-b-fit, message=FALSE, warning=FALSE}
preproc_b          <- preProcess(task1_train %>% select(all_of(lr2_features)),
                                 method = c("center", "scale"))
train_svm_b_scaled <- predict(preproc_b, task1_train %>% select(all_of(lr2_features)))
test_svm_b_scaled  <- predict(preproc_b, task1_test  %>% select(all_of(lr2_features)))

model_svm_b <- svm(
  x           = train_svm_b_scaled,
  y           = task1_y_train,
  kernel      = "radial",
  probability = TRUE
)
```

```{r task1-svm-b-eval, message=FALSE, warning=FALSE}
prob_svm_b <- attr(
  predict(model_svm_b, test_svm_b_scaled, probability = TRUE),
  "probabilities"
)[, "high_tc"]

svm_b_metrics_05 <- evaluate_clf(task1_y_test, prob_svm_b, threshold = 0.5) %>%
  mutate(threshold_type = "0.5")

youden_svm_b <- get_youden_cutoff(task1_y_test, prob_svm_b)

svm_b_metrics_youden <- evaluate_clf(task1_y_test, prob_svm_b, threshold = youden_svm_b) %>%
  mutate(threshold_type = "Youden")

svm_b_metrics <- bind_rows(svm_b_metrics_05, svm_b_metrics_youden) %>%
  mutate(model = "SVM-B (9 feat)") %>%
  dplyr::select(model, threshold_type, everything())

svm_b_metrics
```

```{r task1-svm-b-roc, message=FALSE, warning=FALSE, fig.width=6, fig.height=5}
roc_svm_b <- pROC::roc(task1_y_test, prob_svm_b,
                       levels = c("non_high_tc", "high_tc"), quiet = TRUE)
plot(roc_svm_b,
     main = paste0("ROC — SVM-B (9 features)  AUC = ", round(auc(roc_svm_b), 3)))
youden_coords_svm_b <- pROC::coords(roc_svm_b, x = "best", best.method = "youden",
                                    ret = c("specificity", "sensitivity"), transpose = FALSE)
points(youden_coords_svm_b$specificity, youden_coords_svm_b$sensitivity,
       pch = 19, col = "red", cex = 1.5)
legend("bottomright", legend = "Youden threshold", col = "red", pch = 19, bty = "n")
```

With collinearity removed, SVM-B has a cleaner margin in the kernel space. This is the opposite of random forest or decision trees, where more features help — here, removing redundant dimensions improves performance.

#### Comparison of SVM variants

```{r task1-svm-comparison, message=FALSE, warning=FALSE}
svm_comparison <- bind_rows(svm_a_metrics, svm_b_metrics) %>%
  dplyr::select(model, threshold_type, threshold, accuracy, sensitivity,
                specificity, precision, f1, balanced_acc, auc)

svm_comparison
```

SVM is the least interpretable model evaluated — the RBF kernel maps data into an implicit high-dimensional space and there are no feature coefficients to check. Unlike LR, where collinearity inflates standard errors, in SVM collinearity distorts the margins geometry directly. SVM-B with 9 deduplicated features therefore slightly outperforms SVM-A in AUC (0.926 vs 0.916) and other metrics, confirming that reducing redundancy benefits margin-based classifiers.

### Random Forest

Random forest builds an ensemble method that consists of decision trees, each trained on a bootstrap sample of the data with a random subset of features considered at each split greedily. This randomisation makes it robust to collinearity — correlated features share the splits across trees rather than distorting a single boundary. The full set of 81 numerical predictors is used, allowing the forest to use weak signals not captured by EDA screening that might proove useful in combinaton with other features.

Feature importance is measured by mean decrease in Gini impurity (MeanDecreaseGini), which reflects how much each predictor reduces node impurity across all trees.

```{r task1-rf-fit, message=FALSE, warning=FALSE}
model_rf <- randomForest(
  tc_class ~ .,
  data       = bind_cols(task1_train %>% select(all_of(numeric_predictors)),
                         tc_class = task1_y_train),
  ntree      = 300,
  mtry       = floor(sqrt(length(numeric_predictors))),
  importance = TRUE
)

cat("RF OOB error:", round(model_rf$err.rate[300, "OOB"], 4), "\n")
```

```{r task1-rf-oob, message=FALSE, warning=FALSE, fig.width=7, fig.height=4}
data.frame(
  trees       = seq_len(nrow(model_rf$err.rate)),
  OOB         = model_rf$err.rate[, "OOB"],
  high_tc     = model_rf$err.rate[, "high_tc"],
  non_high_tc = model_rf$err.rate[, "non_high_tc"]
) %>%
  pivot_longer(-trees, names_to = "type", values_to = "error") %>%
  ggplot(aes(trees, error, color = type)) +
  geom_line(linewidth = 0.8) +
  scale_color_manual(
    values = c("OOB" = "black", "high_tc" = "#27ae60", "non_high_tc" = "#e74c3c"),
    name = NULL
  ) +
  labs(title = "OOB error stabilises as more trees are added",
       x = "Number of trees", y = "Error rate") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")
```

```{r task1-rf-eval, message=FALSE, warning=FALSE}
prob_rf <- predict(model_rf,
                   newdata = task1_test %>% select(all_of(numeric_predictors)),
                   type = "prob")[, "high_tc"]

rf_metrics_05 <- evaluate_clf(task1_y_test, prob_rf, threshold = 0.5) %>%
  mutate(threshold_type = "0.5")

youden_rf <- get_youden_cutoff(task1_y_test, prob_rf)

rf_metrics_youden <- evaluate_clf(task1_y_test, prob_rf, threshold = youden_rf) %>%
  mutate(threshold_type = "Youden")

rf_metrics <- bind_rows(rf_metrics_05, rf_metrics_youden) %>%
  mutate(model = "RF (81 feat)") %>%
  dplyr::select(model, threshold_type, everything())

rf_metrics
```

```{r task1-rf-roc, message=FALSE, warning=FALSE, fig.width=6, fig.height=5}
roc_rf <- pROC::roc(task1_y_test, prob_rf,
                    levels = c("non_high_tc", "high_tc"), quiet = TRUE)
plot(roc_rf,
     main = paste0("ROC — Random Forest (81 features)  AUC = ", round(auc(roc_rf), 3)))
youden_coords_rf <- pROC::coords(roc_rf, x = "best", best.method = "youden",
                                 ret = c("specificity", "sensitivity"), transpose = FALSE)
points(youden_coords_rf$specificity, youden_coords_rf$sensitivity,
       pch = 19, col = "red", cex = 1.5)
legend("bottomright", legend = "Youden threshold", col = "red", pch = 19, bty = "n")
```

```{r task1-rf-importance, message=FALSE, warning=FALSE, fig.width=7, fig.height=6}
imp_df <- importance(model_rf) %>%
  as.data.frame() %>%
  rownames_to_column("feature") %>%
  arrange(desc(MeanDecreaseGini)) %>%
  slice_head(n = 20)

ggplot(imp_df, aes(x = reorder(feature, MeanDecreaseGini), y = MeanDecreaseGini)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(title = "Random Forest — Top 20 features by MeanDecreaseGini",
       x = NULL, y = "MeanDecreaseGini") +
  theme_minimal()
```

Random forest achieved the highest AUC (0.978) of all evaluated models, with the Youden threshold giving sensitivity of 0.953 and specificity of 0.934. With 300 trees and random feature subsampling at each split, the ensemble handles correlated predictors itself — unlike LR and SVM where collinearity is harmful, more features help RF by providing more signal per split.

RF with the Youden threshold is the best overall model. Compared to LR-B, which achieves the highest raw sensitivity (0.970), RF misses only
13 more superconductors (37 vs 24 false negatives), but generates approximately 550 fewer false positives. In a laboratory screening context this trade-off strongly
favours RF — 500 fewer unnecessary experiments is a substantial saving, while 18 additional missed candidates is probablly a smaller cost.

The model is partially interpretable through feature importance, but no directional effect can be read from importance scores alone: a high MeanDecreaseGini only
tells us a feature is useful, not whether higher values increase or decrease the probability of `high_tc`.

### Decision Tree

A single decision tree recursively greedely like RF splits the data on the most informative feature at each node. It is the most directly interpretable model 
in this scenario — the tree diagram shows the exact decision path for any prediction, with explicit thresholds and feature names at every split.
The tree is grown with `cp = 0.001` and then pruned back to the complexity parameter that minimises cross-validated error.

```{r task1-dt-fit, message=FALSE, warning=FALSE}
model_dt <- rpart(
  tc_class ~ .,
  data    = bind_cols(task1_train %>% select(all_of(numeric_predictors)),
                      tc_class = task1_y_train),
  method  = "class",
  control = rpart.control(cp = 0.001)
)

best_cp       <- model_dt$cptable[which.min(model_dt$cptable[, "xerror"]), "CP"]
model_dt_pruned <- prune(model_dt, cp = best_cp)
cat("Best cp:", round(best_cp, 6),
    "| Terminal nodes:", model_dt_pruned$numresp, "\n")
```

```{r task1-dt-plot, message=FALSE, warning=FALSE, fig.width=9, fig.height=6}
model_dt_viz <- prune(model_dt, cp = 0.005)
rpart.plot(model_dt_viz, type = 4, extra = 104,
           main = "Decision Tree (simplified view, cp = 0.005)")
```

```{r task1-dt-eval, message=FALSE, warning=FALSE}
prob_dt <- predict(model_dt_pruned,
                   newdata = task1_test %>% select(all_of(numeric_predictors)),
                   type = "prob")[, "high_tc"]

dt_metrics_05 <- evaluate_clf(task1_y_test, prob_dt, threshold = 0.5) %>%
  mutate(threshold_type = "0.5")

youden_dt <- get_youden_cutoff(task1_y_test, prob_dt)

dt_metrics_youden <- evaluate_clf(task1_y_test, prob_dt, threshold = youden_dt) %>%
  mutate(threshold_type = "Youden")

dt_metrics <- bind_rows(dt_metrics_05, dt_metrics_youden) %>%
  mutate(model = "DT (81 feat)") %>%
  dplyr::select(model, threshold_type, everything())

dt_metrics
```

```{r task1-dt-roc, message=FALSE, warning=FALSE, fig.width=6, fig.height=5}
roc_dt <- pROC::roc(task1_y_test, prob_dt,
                    levels = c("non_high_tc", "high_tc"), quiet = TRUE)
plot(roc_dt,
     main = paste0("ROC — Decision Tree (81 features)  AUC = ", round(auc(roc_dt), 3)))
youden_coords_dt <- pROC::coords(roc_dt, x = "best", best.method = "youden",
                                 ret = c("specificity", "sensitivity"), transpose = FALSE)
points(youden_coords_dt$specificity, youden_coords_dt$sensitivity,
       pch = 19, col = "red", cex = 1.5)
legend("bottomright", legend = "Youden threshold", col = "red", pch = 19, bty = "n")
```

The decision tree achieved an AUC of 0.955, placing it between LR/SVM and RF. It is the second most interpretable model after LR — the tree diagram provides a fully
readable decision path. Unlike LR, no p-values are available, but the splits themselves are indication of feature importance. Collinearity is not a concern for
DT since each split greedily picks one feature, but a single tree is more prone to overfitting than a forest — the 0.023 AUC gap versus RF reflects that difference perfectly.

### Overall comparison

```{r task1-overall-comparison, message=FALSE, warning=FALSE}
task1_summary <- bind_rows(
  lr_a_metrics, lr_b_metrics,
  svm_a_metrics, svm_b_metrics,
  rf_metrics,
  dt_metrics
) %>%
  dplyr::select(model, threshold_type, accuracy, sensitivity, specificity,
                f1, balanced_acc, auc, tp, fn, fp)

task1_summary
```

RF with the Youden threshold is the recommended model — it achieves the highest AUC (0.978), best balanced accuracy, and generates far fewer false positives than LR, 
making it most practical for laboratory screening. LR-B with the Youden threshold achieves the highest sensitivity (0.970) and fewest missed superconductors 
(24 false negatives), but this comes at the cost of a very high false positive count (550 more), meaning many unnecessary experiments.
For most real-world screening workflows, the we think this trade-off favours RF saving lots of time and resources.

## Scenario 3: Feature Selection Methods

In the third scenario, we focus on feature selection for the binary classification task. The goal is to compare one algorithmic feature-selection method with two embedded methods. We use the same target variable as in the previous scenario, where superconductors are classified as `high_tc` if their critical temperature is at least 77 K and `non_high_tc` otherwise.

The motivation for this scenario comes from the exploratory data analysis, where several predictors showed strong class-separation ability, but many of them were also highly correlated. Therefore, feature selection is important not only for predictive performance, but also for reducing redundancy and improving model interpretability.

The compared feature-selection approaches are:

-   backward stepwise logistic regression using AIC,
-   lasso logistic regression,
-   elastic net logistic regression.

```{r task3-setup, message=FALSE, warning=FALSE}
load("prepared_data.RData")

candidate_features <- numeric_predictors

train_task3 <- train_df %>%
  dplyr::select(tc_class, all_of(candidate_features))

test_task3 <- test_df %>%
  dplyr::select(tc_class, all_of(candidate_features))

y_train <- if_else(train_task3$tc_class == "high_tc", 1, 0)
y_test  <- if_else(test_task3$tc_class == "high_tc", 1, 0)

x_train <- model.matrix(tc_class ~ ., data = train_task3)[, -1]
x_test  <- model.matrix(tc_class ~ ., data = test_task3)[, -1]
```

The full set of numerical predictors was used as the initial candidate feature space. This allows the feature-selection methods to operate on the complete high-dimensional representation rather than on a manually pre-filtered subset.

```{r task3-helper-functions, message=FALSE, warning=FALSE}
# Helper function: evaluate classification performance.
evaluate_classifier <- function(truth, prob, threshold = 0.5) {
  # Convert predicted probabilities to class labels using a selected threshold.
  pred <- if_else(prob >= threshold, "high_tc", "non_high_tc") %>%
    factor(levels = c("non_high_tc", "high_tc"))
  
  # Ensure that the true labels have the correct factor levels.
  truth <- factor(truth, levels = c("non_high_tc", "high_tc"))
  
  # Compute confusion matrix metrics.
  cm <- confusionMatrix(pred, truth, positive = "high_tc")
  
  # Compute ROC curve and AUC.
  roc_obj <- pROC::roc(
    response = truth,
    predictor = prob,
    levels = c("non_high_tc", "high_tc"),
    quiet = TRUE
  )
  
  # Return relevant metrics in one tibble.
  tibble(
    threshold = threshold,
    accuracy = unname(cm$overall["Accuracy"]),
    sensitivity = unname(cm$byClass["Sensitivity"]),
    specificity = unname(cm$byClass["Specificity"]),
    precision = unname(cm$byClass["Precision"]),
    recall = unname(cm$byClass["Recall"]),
    f1 = unname(cm$byClass["F1"]),
    auc = as.numeric(pROC::auc(roc_obj))
  )
}

# Helper function: compute Youden threshold.
get_youden_threshold <- function(truth, prob) {
  # Ensure that the true labels have the correct factor levels.
  truth <- factor(truth, levels = c("non_high_tc", "high_tc"))
  
  # Build ROC curve.
  roc_obj <- pROC::roc(
    response = truth,
    predictor = prob,
    levels = c("non_high_tc", "high_tc"),
    quiet = TRUE
  )
  
  # Select the threshold that maximizes Youden's index:
  # sensitivity + specificity - 1.
  pROC::coords(
    roc_obj,
    x = "best",
    best.method = "youden",
    ret = "threshold",
    transpose = FALSE
  ) %>%
    as.numeric() %>%
    first()
}
```

### Backward stepwise logistic regression

As the algorithmic feature-selection method, we used backward stepwise logistic regression. The initial model contained all numerical predictors from the prepared dataset. Backward selection starts with this full model and iteratively removes predictors according to the AIC criterion. AIC balances model fit and model complexity, therefore the method should remove predictors that do not sufficiently improve the model.

```{r task3-backward, message=FALSE, warning=FALSE}
full_formula <- as.formula(
  paste("tc_class ~", paste(candidate_features, collapse = " + "))
)

lr_full <- glm(
  full_formula,
  data = train_task3,
  family = binomial()
)

lr_backward <- MASS::stepAIC(
  lr_full,
  direction = "backward",
  trace = FALSE
)

backward_features <- names(coef(lr_backward)) %>%
  setdiff("(Intercept)")

length(backward_features)
```

The final backward-selected model retained `r length(backward_features)` predictors out of the original `r length(candidate_features)`. This means that the method reduced the feature space only moderately. Although some predictors were removed, the final model still remained relatively complex, which suggests that many descriptors contribute to the classification task or that the signal is distributed across correlated predictors.

```{r task3-backward-evaluation, message=FALSE, warning=FALSE}
prob_backward <- predict(
  lr_backward,
  newdata = test_task3,
  type = "response"
)

backward_metrics_05 <- evaluate_classifier(
  truth = test_task3$tc_class,
  prob = prob_backward,
  threshold = 0.5
) %>%
  mutate(threshold_type = "0.5")

backward_youden_threshold <- get_youden_threshold(
  truth = test_task3$tc_class,
  prob = prob_backward
)

backward_metrics_youden <- evaluate_classifier(
  truth = test_task3$tc_class,
  prob = prob_backward,
  threshold = backward_youden_threshold
) %>%
  mutate(threshold_type = "Youden")

backward_metrics <- bind_rows(
  backward_metrics_05,
  backward_metrics_youden
) %>%
  dplyr::select(
    threshold_type,
    threshold,
    accuracy,
    sensitivity,
    specificity,
    precision,
    recall,
    f1,
    auc
  )

backward_metrics
```

The model was evaluated using both the default threshold of 0.5 and the Youden threshold. The Youden threshold was included because the target classes are imbalanced and the minority `high_tc` class is especially important.

The backward-selected logistic regression model achieved an AUC of 0.952, indicating strong overall ability to separate `high_tc` and `non_high_tc` samples. With the default threshold of 0.5, the model reached an accuracy of 0.895, sensitivity of 0.728 and F1-score of 0.731.

The Youden threshold selected from the ROC curve was 0.253. Using this lower threshold substantially increased sensitivity from 0.728 to 0.929, which means that the model detected a much larger proportion of true `high_tc` samples. This improvement came at the cost of lower specificity and precision. Specificity decreased from 0.936 to 0.870 and precision decreased from 0.733 to 0.634. The F1-score increased from 0.731 to 0.753, suggesting a better balance between precision and recall under the Youden threshold.

Overall, backward stepwise logistic regression provided strong predictive performance, but it retained a large number of predictors. Therefore, while the model is useful as a performance-oriented baseline for feature selection, its interpretability remains limited.

### Lasso logistic regression

Lasso logistic regression was used as the first embedded feature-selection method. In `glmnet`, lasso corresponds to `alpha = 1`. This method can shrink some coefficients exactly to zero, which means that it performs feature selection directly during model fitting.

```{r task3-lasso-fit, message=FALSE, warning=FALSE}
set.seed(42)

cv_lasso <- cv.glmnet(
  x = x_train,
  y = y_train,
  family = "binomial",
  alpha = 1,
  type.measure = "auc"
)
```

The cross-validation criterion was AUC because the dataset is imbalanced and accuracy alone may be misleading. Two lambda values were evaluated:

-   `lambda.min`, which gives the best cross-validated AUC,
-   `lambda.1se`, which gives a simpler model within one standard error of the best model.

```{r task3-lasso-lambda-min, message=FALSE, warning=FALSE}
lasso_coef_min <- coef(cv_lasso, s = "lambda.min")

lasso_features_min <- rownames(lasso_coef_min)[as.vector(lasso_coef_min != 0)] %>%
  setdiff("(Intercept)")

prob_lasso_min <- predict(
  cv_lasso,
  newx = x_test,
  s = "lambda.min",
  type = "response"
) %>%
  as.vector()

lasso_metrics_min_05 <- evaluate_classifier(
  truth = test_task3$tc_class,
  prob = prob_lasso_min,
  threshold = 0.5
)

lasso_youden_threshold_min <- get_youden_threshold(
  truth = test_task3$tc_class,
  prob = prob_lasso_min
)

lasso_metrics_min_youden <- evaluate_classifier(
  truth = test_task3$tc_class,
  prob = prob_lasso_min,
  threshold = lasso_youden_threshold_min
)

lasso_metrics_min <- bind_rows(
  lasso_metrics_min_05 %>%
    mutate(lambda_type = "lambda.min", threshold_type = "0.5"),
  lasso_metrics_min_youden %>%
    mutate(lambda_type = "lambda.min", threshold_type = "Youden")
) %>%
  dplyr::select(lambda_type, threshold_type, everything())

length(lasso_features_min)
lasso_metrics_min
```

```{r task3-lasso-lambda-1se, message=FALSE, warning=FALSE}
lasso_coef_1se <- coef(cv_lasso, s = "lambda.1se")

lasso_features_1se <- rownames(lasso_coef_1se)[as.vector(lasso_coef_1se != 0)] %>%
  setdiff("(Intercept)")

prob_lasso_1se <- predict(
  cv_lasso,
  newx = x_test,
  s = "lambda.1se",
  type = "response"
) %>%
  as.vector()

lasso_metrics_1se_05 <- evaluate_classifier(
  truth = test_task3$tc_class,
  prob = prob_lasso_1se,
  threshold = 0.5
)

lasso_youden_threshold_1se <- get_youden_threshold(
  truth = test_task3$tc_class,
  prob = prob_lasso_1se
)

lasso_metrics_1se_youden <- evaluate_classifier(
  truth = test_task3$tc_class,
  prob = prob_lasso_1se,
  threshold = lasso_youden_threshold_1se
)

lasso_metrics_1se <- bind_rows(
  lasso_metrics_1se_05 %>%
    mutate(lambda_type = "lambda.1se", threshold_type = "0.5"),
  lasso_metrics_1se_youden %>%
    mutate(lambda_type = "lambda.1se", threshold_type = "Youden")
) %>%
  dplyr::select(lambda_type, threshold_type, everything())

length(lasso_features_1se)
lasso_metrics_1se
```

```{r task3-lasso-summary, message=FALSE, warning=FALSE}
lasso_summary <- bind_rows(
  lasso_metrics_min %>%
    mutate(
      lambda_type = "lambda.min",
      retained_features = length(lasso_features_min)
    ),
  lasso_metrics_1se %>%
    mutate(
      lambda_type = "lambda.1se",
      retained_features = length(lasso_features_1se)
    )
) %>%
  dplyr::select(
    lambda_type,
    threshold_type,
    retained_features,
    threshold,
    accuracy,
    sensitivity,
    specificity,
    precision,
    recall,
    f1,
    auc
  )

lasso_summary
```

### Elastic net logistic regression

Elastic net was used as the second embedded feature-selection method. It combines lasso and ridge penalties. The `alpha` parameter controls this combination: values closer to 1 behave more like lasso, while values closer to 0 behave more like ridge.

```{r task3-elastic-net-fit, message=FALSE, warning=FALSE}
set.seed(42)

alpha_grid <- c(0.2, 0.5, 0.8)

enet_cv_results <- tibble(alpha = alpha_grid) %>%
  mutate(
    cv_fit = purrr::map(
      alpha,
      ~ cv.glmnet(
        x = x_train,
        y = y_train,
        family = "binomial",
        alpha = .x,
        type.measure = "auc"
      )
    ),
    best_auc = purrr::map_dbl(cv_fit, ~ max(.x$cvm))
  ) %>%
  arrange(desc(best_auc))

enet_cv_results
```

The elastic net model with the highest cross-validated AUC was selected for final evaluation.

```{r task3-elastic-net-selection, message=FALSE, warning=FALSE}
best_enet <- enet_cv_results$cv_fit[[1]]
best_alpha <- enet_cv_results$alpha[[1]]

best_alpha
```

```{r task3-elastic-net-lambda-min, message=FALSE, warning=FALSE}
enet_coef_min <- coef(best_enet, s = "lambda.min")

enet_features_min <- rownames(enet_coef_min)[as.vector(enet_coef_min != 0)] %>%
  setdiff("(Intercept)")

prob_enet_min <- predict(
  best_enet,
  newx = x_test,
  s = "lambda.min",
  type = "response"
) %>%
  as.vector()

enet_metrics_min_05 <- evaluate_classifier(
  truth = test_task3$tc_class,
  prob = prob_enet_min,
  threshold = 0.5
) %>%
  mutate(
    lambda_type = "lambda.min",
    threshold_type = "0.5",
    retained_features = length(enet_features_min)
  )

enet_youden_threshold_min <- get_youden_threshold(
  truth = test_task3$tc_class,
  prob = prob_enet_min
)

enet_metrics_min_youden <- evaluate_classifier(
  truth = test_task3$tc_class,
  prob = prob_enet_min,
  threshold = enet_youden_threshold_min
) %>%
  mutate(
    lambda_type = "lambda.min",
    threshold_type = "Youden",
    retained_features = length(enet_features_min)
  )

length(enet_features_min)
```

```{r task3-elastic-net-lambda-1se, message=FALSE, warning=FALSE}
enet_coef_1se <- coef(best_enet, s = "lambda.1se")

enet_features_1se <- rownames(enet_coef_1se)[as.vector(enet_coef_1se != 0)] %>%
  setdiff("(Intercept)")

prob_enet_1se <- predict(
  best_enet,
  newx = x_test,
  s = "lambda.1se",
  type = "response"
) %>%
  as.vector()

enet_metrics_1se_05 <- evaluate_classifier(
  truth = test_task3$tc_class,
  prob = prob_enet_1se,
  threshold = 0.5
) %>%
  mutate(
    lambda_type = "lambda.1se",
    threshold_type = "0.5",
    retained_features = length(enet_features_1se)
  )

enet_youden_threshold_1se <- get_youden_threshold(
  truth = test_task3$tc_class,
  prob = prob_enet_1se
)

enet_metrics_1se_youden <- evaluate_classifier(
  truth = test_task3$tc_class,
  prob = prob_enet_1se,
  threshold = enet_youden_threshold_1se
) %>%
  mutate(
    lambda_type = "lambda.1se",
    threshold_type = "Youden",
    retained_features = length(enet_features_1se)
  )

length(enet_features_1se)
```

```{r task3-elastic-net-summary, message=FALSE, warning=FALSE}
enet_summary <- bind_rows(
  enet_metrics_min_05,
  enet_metrics_min_youden,
  enet_metrics_1se_05,
  enet_metrics_1se_youden
) %>%
  mutate(
    method = paste0("Elastic Net LR (alpha=", best_alpha, ")")
  ) %>%
  dplyr::select(
    method,
    lambda_type,
    threshold_type,
    retained_features,
    threshold,
    accuracy,
    sensitivity,
    specificity,
    precision,
    recall,
    f1,
    auc
  )

enet_summary
```

### Overall comparison of feature-selection methods

The following table compares the feature-selection methods in terms of predictive performance and the number of retained predictors.

```{r task3-final-summary, message=FALSE, warning=FALSE}
task3_summary <- bind_rows(
  backward_metrics %>%
    mutate(
      method = "Backward LR",
      lambda_type = "AIC",
      retained_features = length(backward_features)
    ) %>%
    dplyr::select(
      method,
      lambda_type,
      threshold_type,
      retained_features,
      threshold,
      accuracy,
      sensitivity,
      specificity,
      precision,
      recall,
      f1,
      auc
    ),
  lasso_summary %>%
    mutate(
      method = "Lasso LR"
    ) %>%
    dplyr::select(
      method,
      lambda_type,
      threshold_type,
      retained_features,
      threshold,
      accuracy,
      sensitivity,
      specificity,
      precision,
      recall,
      f1,
      auc
    ),
  enet_summary
)

task3_summary
```

The results show that the feature-selection methods reduced the original predictor space only moderately. Backward stepwise logistic regression retained a relatively large number of predictors, while lasso with `lambda.1se` produced the most compact standard model. However, the performance differences between the less sparse and more sparse models were relatively small, especially when using the Youden threshold.

### Feature retention comparison

In addition to predictive performance, we compared which features were retained by each feature-selection method. Features selected by several methods can be considered more stable candidates.

```{r task3-feature-retention, message=FALSE, warning=FALSE}
feature_retention <- tibble(feature = candidate_features) %>%
  mutate(
    backward = feature %in% backward_features,
    lasso_min = feature %in% lasso_features_min,
    lasso_1se = feature %in% lasso_features_1se,
    elastic_net_min = feature %in% enet_features_min,
    elastic_net_1se = feature %in% enet_features_1se,
    selected_count = backward + lasso_min + lasso_1se + elastic_net_min + elastic_net_1se
  ) %>%
  arrange(desc(selected_count), feature)

feature_retention
```

```{r task3-stable-features, message=FALSE, warning=FALSE}
stable_features <- feature_retention %>%
  filter(selected_count == 5) %>%
  arrange(desc(selected_count), feature)

stable_features
```

The table above lists features that were selected by all five evaluated variants. These features can be interpreted as the most stable predictors across the tested feature-selection methods.

```{r task3-feature-selection-counts, message=FALSE, warning=FALSE}
feature_selection_counts <- feature_retention %>%
  summarise(
    backward = sum(backward),
    lasso_min = sum(lasso_min),
    lasso_1se = sum(lasso_1se),
    elastic_net_min = sum(elastic_net_min),
    elastic_net_1se = sum(elastic_net_1se)
  ) %>%
  pivot_longer(
    cols = everything(),
    names_to = "method",
    values_to = "retained_features"
  ) %>%
  arrange(retained_features)

feature_selection_counts
```

Overall, the scenario shows that the predictive signal is not concentrated in only a small number of predictors. Instead, many material descriptors contribute to the classification task. Regularized methods, especially lasso with `lambda.1se`, provide a better trade-off between model complexity and predictive performance, while backward stepwise regression keeps more predictors and is therefore less interpretable.

# 15. Final Model Selection

Based on the results, the final model should be selected according to both performance and interpretability.

The selected model should provide a good compromise between classification performance, number of retained features, and explainability.

# 16. Limitations

The main limitations of this project are:

-   the original regression problem was simplified into binary classification,
-   the threshold of 77 K is practically meaningful but still simplifies the target variable,
-   class imbalance may influence model evaluation,
-   correlated predictors can affect feature-selection results,
-   the models identify statistical relationships, not direct physical causality.

# 17. Conclusion

This project compared several classification and feature-selection approaches for identifying high-temperature superconductors. The results show whether regularized models can reduce the number of input features while maintaining good predictive performance.

# 18. Reproducibility

```{r}
sessionInfo()
```
